Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-06-23 draft_lm1.tex
Removed m from equation; closes #39
added response letter; added some responses to the manuscript. changes the code which generates the table in the paper.

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate} and $S_0$ is the initial fraction of susceptible 
individuals.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

@@

We take a Bayesian approach to the estimation of $S_0$.
First the incidence time series was divided into $J=13$ 
epidemic windows that corresponded to significant raises in incidence and 
normalized to lie on the $[0,1]$ interval.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, m, \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is assumed to be a Normal distribution 
with fixed variance $\sigma^2$.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described above and fixed $\tau = 1/7\: \text{days}^{-1}$.
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. 
Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic, however the peak of cases
may occur in the following year.
$^\ddag$: Susceptible fraction.
These results show considerable variation in AR between epidemics, consistent with
the accquiring and loss of serotype-specific immunity.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & median Attack Ratio & $S_0^\ddag$ \\
\hline
1996 & 0.39 (0.17-0.54) & 0.00171(0.0012-0.0038)\\
1997 & 0.87 (0.74-0.87) & 0.00273(0.0027-0.0032)\\
1998 & 0.5 (0.49-0.5) & 0.00142(0.0014-0.0014)\\
1999 & 0.11 (0.037-0.2) & 0.00345(0.0018-0.01)\\
2000 & 0.25 (0.24-0.27) & 0.0155(0.015-0.016)\\
2001 & 0.48 (0.47-0.49) & 0.0495(0.048-0.051)\\
2005 & 0.15 (0.1-0.21) & 0.0147(0.01-0.021)\\
2006 & 0.11 (0.08-0.14) & 0.0281(0.022-0.037)\\
2007 & 0.15 (0.15-0.15) & 0.135(0.13-0.14)\\
2008 & 0.14 (0.031-0.31) & 0.00672(0.003-0.024)\\
2010 & 0.18 (0.17-0.19) & 0.0454(0.043-0.048)\\
2011 & 0.086 (0.082-0.094) & 0.215(0.2-0.23)\\
2012 & 0.14 (0.13-0.15) & 0.0621(0.058-0.068)\\
\hline
\end{tabular}
\end{center}
\label{tab:AR}
\end{table}

>>>>

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate} and $S_0$ is the initial fraction of susceptible 
individuals.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations. Although demography is not included in the model -- population is 
assumed to remain constant throughout an epidemic -- the population variation 
from year to year is taken into consideration as the prevalences are calculated 
for each year by dividing the number of cases by the official population 
estimated by the census. In Brazil, dengue affects all age groups which still 
haven't been exposed to all 4 serotypes. This is in contrast with countries in 
which the four serotypes have been endemic for a very long period of time, 
where dengue mostly affect the youth.

Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

@@

We take a Bayesian approach to the estimation of $S_0$.
First the incidence time series was divided into $J=13$ 
epidemic windows that corresponded to significant raises in incidence and 
normalized to lie on the $[0,1]$ interval.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is assumed to be a Normal distribution 
with fixed variance $\sigma^2$.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described above and fixed $\tau = 1/7\: \text{days}^{-1}$.
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. 
Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic, but the peak of cases
may occur in the following year.
$^\ddag$: percentage of population.
These results show considerable variation in AR between epidemics, consistent with
the accquiring and loss of serotype-specific immunity.}
\begin{center}
\begin{tabular}{l|c|c|c}
\hline
$^\dag$Year & $^\ddag$Cases & median Attack Ratio & $^\ddag$$S_0$ \\
\hline
1996 & 0.066 & 0.39 (0.17-0.54) & 0.171(0.12-0.38)\\
1997 & 0.238 & 0.87 (0.74-0.87) & 0.273(0.27-0.32)\\
1998 & 0.0708 & 0.5 (0.49-0.5) & 0.142(0.14-0.14)\\
1999 & 0.0371 & 0.11 (0.037-0.2) & 0.345(0.18-1.0)\\
2000 & 0.394 & 0.25 (0.24-0.27) & 1.55(1.5-1.6)\\
2001 & 2.38 & 0.48 (0.47-0.49) & 4.95(4.8-5.1)\\
2005 & 0.217 & 0.15 (0.1-0.21) & 1.47(1.0-2.1)\\
2006 & 0.315 & 0.11 (0.08-0.14) & 2.81(2.2-3.7)\\
2007 & 2.03 & 0.15 (0.15-0.15) & 13.5(1.3e+01-1.4e+01)\\
2008 & 0.0923 & 0.14 (0.031-0.31) & 0.672(0.3-2.4)\\
2010 & 0.831 & 0.18 (0.17-0.19) & 4.54(4.3-4.8)\\
2011 & 1.85 & 0.086 (0.082-0.094) & 21.5(2e+01-2.3e+01)\\
2012 & 0.864 & 0.14 (0.13-0.15) & 6.21(5.8-6.8)\\

\hline
\end{tabular}
\end{center}
\label{tab:AR}
\end{table}
